---
title: "The most important skill to a data scientist is email"
author: "Havis, Koon, Tillmawitz"
format: html
---

```{r}
library(tidyverse)
library(treemap)
```


## Introduction

## ONET

The O*NET database includes information on skills, abilities, knowledge, work activities, and interests associated with occupations. This information can be used to facilitate career exploration, vocational counseling, and a variety of human resources functions, such as developing job orders and position descriptions and aligning training with current workplace needs.

```{r}
# Load all ONET data
source('../data/raw/load_onet_historical_data.R')
```
### Soft skills

For our analysis we have isolated Data Scientists and Statisticians as the primary occupations of interest. Note that Data Scientist is considered a "parent" occupation, and includes the sub-occupations of Business Intelligence Analyst and Clinical Data Managers. O\*NET groups occupations according to how similar it finds their "work activities" to be.

```{r}
skills <- onet_24$skills

kable(head(skills))
```



```{r}
top_skills <- skills |> 
  group_by(element_name, scale_id, data_value, .drop = TRUE) |> 
  summarize(data_value = mean(data_value), .groups = "drop") |> 
  arrange(desc(data_value)) 

ggplot(top_skills, aes(x = fct_reorder(element_name, data_value, .desc = FALSE), y = data_value, fill = scale_id)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  theme_minimal()
```


```{r}
abilities <- onet_24$abilities
```



### Technology skills


```{r}
tech_skills <- onet_24$tech

kable(head(tech_skills))
```

"Hot Technology" indicates that this technology skill is popular among all professions, while "in demand" indicates requirements frequently included in job postings for the particular occupation.

```{r}
top_tech <- tech_skills |>  
  filter(in_demand == "Y", soc_name == "data_scientist") |> 
  select(commodity_title, example, hot_technology, in_demand) |> 
  arrange(commodity_title) |> 
  rename("Category" = commodity_title, "Software" = example, "Hot Technology" = hot_technology, "In Demand" = in_demand)

kable(top_tech)
```
